Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-21 paper/main.tex
clean up some bullet points
rewrite overview of package

%###################################################################################################
\section{Overview}
This work consists of many subcomponents which can be combined in order to achieve various tasks.
Figure~\ref{fig:flowchart-project-structure} shows a high-level overview of all components.
We assume that bacterial rods can be represented by a colleciton of vertices (see
section~\ref{subsec:spatial-representation}).
This assumption alone is enough to enable a cascade of other components.
Using the spatial representation, we provide visualization methods that create 3D models from the
given vertices which can then either be rendered or projected to a 2D plane.
The data extraction component determines the vertices from given cell masks which
correspond to microscopic images.
This allows us in particular to turn any sequence of images (movie) into a sequence of bacterial
positions.
The dynamics of the vertices are determined by a computational model which is specific
to the respective use-case and should be defined by the researcher.
It merely needs to utilize the aforementioned spatial representation of bacterial rods as a
collection of vertices.
This work will provide one example of how such a model can be defined.
Notably, the visualization and data extraction components do not require any knowledge about the
computational model, thus allowing us to reuse their functionality across multiple models.

Constructing a computational model requires knowledge about the target system and thus allows us to
interpret the results gathered by the provided data extraction methods.
These results can then further be used to restrict or determine the parameters of our model.
For example, by extracing the positions of bacteria across a time-series of microscopic images, we
can also calculate their lengths at these time-points and thus determine their rate of growth and
type of growth (see section~\ref{subsec:parameter-estimation-individual-treatment}), thus
effectively guiding our modeling steps.

Visualization methods are an integral part of any modeling endevaour.
They allow researchers to obtain feedback quickly and improve iteration times when constructing new
models. Furthermore, we utilize them in our more advanced parameter estimation methods (see
section~\ref{section:parameter-estimation}) where they enable us to effectively compare
two bacterial configurations across division events.
In the future, we are planning to build on the current state and generate realistic microscopic
images and time-series.
Together with the parameter estimation methods, they can then be used to provide large amounts of
mechanistic data for the training of cell-segmentation and cell-tracking machine-learning
algorithms.

\textbf{
    Du musst den Bogen hinbekommen von dem allgemeinen Problem des Parameterschätzen bei ABMs zu dem
    spezifischen Problem, welches du hier verhandelst.
}

\begin{figure}[H]
    \centering
    % \includegraphics[width=0.8\textwidth]{overview.pdf}
    \includestandalone[width=0.8\textwidth]{figures/concept/overview}
    \caption{
        The project can be divided into subcomponents (yellow, red) that interact with and depend on
        each other.
        Yellow blocks are provided by this work and red blocks require additional input such as
        modeling assumptions or data.
        Blue blocks indicate possible future extensions of this project.
        Every component implicitly builds on the spatial representation of rod-shaped bacteria as
        vertices.
    }
    \label{fig:flowchart-project-structure}
\end{figure}

@@

@Christian
\begin{itemize}
    \item shortened everything
    \item leave out parts which are not directly used by the model
    \item move biological context to respective subsections to link more directly to model
    \item remove duplicates and useless text
    \item move heterogeneity considerations from table in main text
\end{itemize}

>>>>

%###################################################################################################
\section{Overview}
This work consists of many subcomponents which can be combined in order to achieve various tasks.
Figure~\ref{fig:flowchart-project-structure} shows a high-level overview of all components.
We assume that bacterial rods can be represented by a colleciton of vertices (see
section~\ref{subsec:spatial-representation}).
This assumption alone is enough to enable a cascade of other components.
Using the spatial representation, we provide visualization methods that employ a 3D rendering
engine and can produce cell masks or rendered images from a given configuration.
The data extraction component determines the vertices from given segmented microscopic images.
This allows us in particular to turn any sequence of images (movie) into a sequence of bacterial
positions and is a key feature of the parameter estimation methods.
The dynamics of the vertices are determined by a computational model which is specific
to the respective use-case and should be defined by the researcher.
It merely requires the aforementioned assumption of representing bacterial rods as a
collection of vertices.
This work will provide one example of how such a model can be defined.
Notably, the visualization and data extraction components do not require any knowledge about the
computational model, thus allowing us to reuse their functionality across differing models.

By utilizing the provided data extraction methods, we can infer properties of the bacteria that aid
us in constructing a suitable computational model.
However, the construction of any mechanistic model also requires prior konwledge which can be used
to inform our analysis methods.
For example, by extracing the positions of bacteria across a time-series of microscopic images, we
can also calculate their lengths at these time-points and thus determine their rate of growth and
type of growth (see section~\ref{subsec:parameter-estimation-individual-treatment}), thus
effectively guiding our modeling steps.
As we will show later, extracted data in the form of cellular positions or segmentation masks can
also be used for comparing computational results with biological ground truth.

Visualization methods are an integral part of any modeling endevaour.
They allow researchers to obtain feedback quickly and improve iteration times when constructing new
models.
Furthermore, we utilize them in our more advanced imaging-based comparison methods (see
section~\ref{subsec:imaging-based-comparison}) where they enable us to effectively compare
two bacterial configurations while accounting for division events.
In the future, we are planning to extend the the rendering capabilities of our approach and
generate realistic microscopic images and time-series.
Together with the parameter estimation methods, they can then be used to provide large amounts of
mechanistic data for the training of cell-segmentation and cell-tracking machine-learning
algorithms.

\begin{figure}[H]
    \centering
    \includestandalone[width=0.8\textwidth]{figures/concept/overview}
    \caption{
        The project can be divided into subcomponents that interact with and depend on
        each other.
        Every component implicitly builds on the spatial representation of rod-shaped bacteria as
        vertices (see section~\ref{subsec:spatial-representation}).
        Red boxes highlight the key scientific drivers of this study while yellow boxes indicate
        auxiliary methods.
        The blue box "Data Generation" displays a extension of this approach which is under active
        development.
    }
    \label{fig:flowchart-project-structure}
\end{figure}

@@
